Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-07 Manuscript/MPESubmission/Revision/response.tex

"In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living or fossil data, or all data for certain characters. % NC: All data overall suggests an empty matrix
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian Inference methods: (i) the conservation of clades and (ii) the displacement of wild-card taxa." lines 99-103 @@@

% "We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred % NC: Didn't one reviewer have issues with the word inferred? TG: for the matrices only (and he/she's right); trees are still inferred.
% using Maximum Likelihood and Bayesian methods by using two different topological distance metrics, based on the Robinson-Foulds (Robinson and Foulds, 1981) and Triplets (Critchlow et al., 1996) distances.
% Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and two distinct topological distance metrics." lines 127-134 @@@
% NC: Why are the distinct topological metrics novel? TG: not novel but novel to use two aspects (people used only clade related metrics before).

% TG: I actually rewrote this bit as well:
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We inferred the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measured the differences in topology using two different topological distance metrics as proxies for clade conservation (based on the Robinson-Foulds distance; Robinson and Foulds, 1981) and for wild-card taxa placement (based on the Triplets distance; Critchlow et al., 1996).
Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring difference on two aspects of topology (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).

@@

We fixed this issue.

% -----------------
% Specific Comment 9
% -----------------
\item{\textcolor{blue}{\textbf{line 76 and throughout:} correct use of the word 'data' from singular to plural throughout}}

% TG: actually I'm not going to change it:
We thank the reviewer for pointing out this inconsistency, however, 'data' is now commonly used for the singular case (Pinker, S. \textit{The Sense of Style: The Thinking Person's Guide to Writing in the 21st Century}. Penguin, 2014. p 271; and Guardian and Observer style guide (2015): \url{http://goo.gl/cmKjNe}).

% NC: Hmmm I reckon this is probably inconsistent throughout then as I usually correct it to plural. I'd just change this tbh. Or at least check you've been consistent. TG: ok maybe be I have a plural/singular problem (I definitely have). I actually never know when to use it a singular or plural: "One data point", that's ok, but "morphological data for one taxa"? Would that be singular or plural?

@@

(1) - We thank the reviewer for this comment and now emphasize both the novel aspects of our study and the questions we are actually testing (see reviewer 1 comment 4 above).

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting. However, we considered this study to already be long and complex. We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might lose the reader and make an already long paper even longer.
Nonetheless, we added the result of these analyses to \textbf{Appendix-A Morphological rates estimations results}.
We link to this analysis in the methods as follows:

"The detailed MrBayes parameters are available in Appendix A along with details on the $\alpha$ parameter estimation." lines 336-338 @@@
See also Appendix A "Morphological rates estimations results".
% TG: what do you think? I don't want to attract the reviewer's attention too much (i.e. that's another shit load of analysis to do to do it properly) and certainly not the reader's attention either (it's not relevant to the whole paper) but sticking that to the supplementary kind of does what the reviewers wants to know.

% NC: I'm a bit confused - did you do what they wanted or not? If not then the whole "this will make the paper longer" argument holds. If you did do what they wanted then it might need a little more explanation somewhere. You can't just throw random stuff into the appendices for no reason.
% TO DISCUSS @@@
% TG: Ok, maybe we should remove it. I've done the analysis because it's interesting and easy to do but maybe we can just add the results to this response, not to the revised manuscript?

@@

% -----------------
% Specific Comment 15
% -----------------
\item{\textcolor{blue}{\textbf{line 125:} omit 'note that we explain each step in detail below this general outline', replace with something like 'Our general protocol included the following steps (Figure 1):'…}}

We removed the part suggested by the reviewer.

% -----------------
% Specific Comment 16
% -----------------
\item{\textcolor{blue}{\textbf{line 167 and throughout manuscript:} replace usage of 'inferred' when referring to simulating data. Replace with 'generated', 'simulated', or 'created', for example. 'Infer' should only be used when a model is fit to data to infer some output such as a tree, transition rates, etc. }}

We replace 'inferred' by 'simulated' when referring to simulating data throughout the manuscript.
% NC: Check this - I'm sure I caught one up above in the comments here. TG: I think the idea is that we didn't "infer" any matrices, we simulated/generated/created them. The trees are still inferred though.

@@

We modified the sentence as suggested.

% -----------------
% Specific Comment 18
% -----------------
\item{\textcolor{blue}{\textbf{line 173 to 176:} need some justification for why the substitution rate with alpha shape of 0.5  will avoid homoplasy. Yang 1996 is cited but clarify the justification of what is 'too much' homoplasy. }}

We changed the sentence to:
"In practice, a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal (Hassanin et al., 1998; Estoup et al., 2002).%\citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below)." (line 189-192@@@).

% NC: I think what is needed here is a definition of what is "too much". Is it 50%? 20%? etc. Too much is arbitrary.
% TO DISCUSS @@@
% TG: That's why I removed the "too much". I'm actually not sure how Homoplasy is measured (as in units of homoplasy). Wagner did a lot of work on it but he usually refers to homoplasy through proxies like characters retention, etc...

@@

% -----------------
% Specific Comment 19
% -----------------
\item{\textcolor{blue}{\textbf{Line 176 to 180:} could be streamlined. It is not clear what is meant by 'no special assumption about how the characters evolved', what constitutes a special assumption? Could potentially be reduced to something like 'this model and these parameter settings strike a balance between realism for empirical datasets (citations) and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations'. Just for example.}}

We modified the sentence as follows:
% "We selected this model and these parameters to strike a balance between realism for empirical datasets (e.g. Kelly et al., 2014) and models with few parameters (cf. Springer et. al., 2012) to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years)." lines 192-198 @@@

"This model and these parameter settings strike a balance between realism for empirical datasets  (e.g. Douady, Kelly et al., 2014) and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years)". lines @@@

% NC: Why not just use the phrasing given by the reviewer for the first part? TG: Done.

@@

\begin{figure}[]
\centering
    \includegraphics[keepaspectratio=true]{response_fig1.pdf}
\caption{Figure 3 from Wright and Hillis 2014 with added x axis ticks. The red line represents the 0.5 rate.}
\end{figure}

We do agree, however, that if rates get lower than 0.5, variance in topological error increases (especially for the Mk model with missing data).
Therefore, we implemented the \textit{a posteriori} matrix selection step to our analysis (i.e. keeping only the ``complete'' matrices that led to trees with a median bootstrap $>$ 50; explained lines 259-268 in the submitted manuscript and lines @@@ and @@@ in the revised one).

We rephrased the section underlined by the reviewer as follows:

"We used low evolutionary rate parameters to be consistent with the molecular rate parameters and to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal (Wright and Hillis, 2014).
In fact, topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model (Lewis, 2001; Wright and Hillis, 2014).
Note, however, that Wright and Hillis (2014) have shown that low morphological rates (< 0.5) increase variance in topological error but we discarded simulations with such topological error by selecting only matrices with a “fair” phylogenetic signal (see Estimating phylogenies section below; Zander, 2004)." lines 211-217 @@@

% NC: Might need a bit more explanation. For example mentioning that "Topological error, however, becomes minimal for all scenarios (Parsimony and Mk, with or without missing data) at a rate of 0.5 (see figure 3 from Wright and Hillis (2014)" as you've mentioned above.
% TG: added a sentence.

### 2015-08-07 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Updated revisions

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
\hl{These approaches differ mainly in how they treat the fossil taxa and it's data.
One can use the fossil as tips or as nodes in the phylogeny and use it's age only, it's morphology only or it's age and morphology jointly.}
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
\hl{Because fossil taxa do not usually have available DNA, only fossils' occurrence dates are used to time calibrate phylogenies} \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the ``best'' approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}.
Neither approach, however, uses all the available data.

@@

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
%in
\hl{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We inferred the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measured the differences in topology using two different topological distance metrics as proxies for clade conservation } \citep[\hl{based on the Robinson-Foulds distance;}][]{RF1981}\hl{ and for wild-card taxa placement }\citep[\hl{based on the Triplets distance;}][]{critchlowthe1996}.
\hl{Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring difference on two aspects of topology }\citep[c.f.][\hl{measuring respectively only the clades support or the number of shared clades}]{Wiens01102005,pattinsonphylogeny2014}.
%Former revised version
% \hl{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods by using two different topological metrics, based on the Robinson-Foulds }\citep{RF1981}\hl{ and Triplets }\citep{critchlowthe1996}\hl{ distances.
% Thus, in the present study, we propose a new approach on the missing-data question focus on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and two distinct topological metrics.}
% We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimizing the amount of missing data in the fossil record.
% Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.

@@

Next, we generated a molecular and a morphological matrix from the ``true'' tree.
The molecular matrix was \hl{simulated} from the ``true'' tree using the R package phyclust v. 0.1-14 \citep{chen2011}.
The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}.
\hl{The substitution rates were selected from a gamma distribution with an ($\alpha$) shape of 0.5} \citep{yangamong-site1996}.
\hl{In practice, such a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal} \citep{Hassanin1998611,EstoupHomoplasy}.
\hl{Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below).}
%We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. %Justify
%We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years).
\hl{We selected this model and parameters settings to be a compromise between realism from empirical datasets} \citep[e.g.][]{kellymolecular2014} \hl{and models with few parameters} \citep[cf.][]{springermacroevolutionary2012} \hl{to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years).}
All the molecular information for fossil taxa was replaced by missing data ("?").

\hl{We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11} \citep{paradisape:2004}\hl{ to generate a matrix of 100 character sites for 51 taxa.}
\hl{We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).}
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}.
This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
\hl{We used low evolutionary rate parameters to be consistent with the molecular rates parameters and to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal} \citep{wrightbayesian2014}.
\hl{Note that} \cite{wrightbayesian2014} \hl{have shown that low morphological rates ($<0.5$) increases variance in topological error but we discarded simulations with such topological error by selecting only matrices with a ``fair'' phylogenetic signal} \citep[\hl{see Estimating phylogenies section below;}][]{zanderminimal2004}.

>>>>

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
\hl{These approaches differ mainly in how they treat fossil taxa and their data.
One can use fossils as tips or as nodes in the phylogeny, and can use only the age of the fossils, only the morphology of the fossils, or age and morphology jointly.}
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
\hl{Because fossil taxa do not usually have available DNA, only fossil occurrence dates are used to time calibrate phylogenies }\citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the ``best'' approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}.
Neither approach, however, uses all the available data.

@@

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
%in
\hl{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We inferred the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measured the differences in topology using two different topological distance metrics as proxies for clade conservation } \citep[\hl{based on the Robinson-Foulds distance;}][]{RF1981}\hl{ and for wild-card taxa placement }\citep[\hl{based on the Triplets distance;}][]{critchlowthe1996}.
\hl{Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring difference on two aspects of topology }\citep[cf.][\hl{measuring respectively only the clades support or the number of shared clades}]{Wiens01102005,pattinsonphylogeny2014}.
%Former revised version
% \hl{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods by using two different topological metrics, based on the Robinson-Foulds }\citep{RF1981}\hl{ and Triplets }\citep{critchlowthe1996}\hl{ distances.
% Thus, in the present study, we propose a new approach on the missing-data question focus on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and two distinct topological metrics.}
% We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimizing the amount of missing data in the fossil record.
% Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.

@@

Next, we generated a molecular and a morphological matrix from the ``true'' tree.
The molecular matrix was \hl{simulated} from the ``true'' tree using the R package phyclust v. 0.1-14 \citep{chen2011}.
The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}.
\hl{The substitution rates were selected from a gamma distribution with an ($\alpha$) shape of 0.5} \citep{yangamong-site1996}.
\hl{In practice, such a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal} \citep{Hassanin1998611,EstoupHomoplasy}.
\hl{Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below).}
%We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. %Justify
%We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years).
\hl{This model and these parameter settings strike a balance between realism for empirical datasets }\citep[e.g.][]{douadycomparison2003,kellymolecular2014}\hl{ and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years).}
All the molecular information for fossil taxa was replaced by missing data ("?").

\hl{We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11} \citep{paradisape:2004}\hl{ to generate a matrix of 100 character sites for 51 taxa.}
\hl{We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).}
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}.
This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
\hl{We used low evolutionary rate parameters to be consistent with the molecular rates parameters and to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal} \citep{wrightbayesian2014}.
\hl{In fact, topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model }\citep{lewisa2001,wrightbayesian2014}.
\hl{Note, however, that } \cite{wrightbayesian2014}\hl{ have shown that low morphological rates (< 0.5) increase variance in topological error but we discarded simulations with such topological error by selecting only matrices with a “fair” phylogenetic signal }\citep[\hl{see Estimating phylogenies section below;}][]{zanderminimal2004}.
